---
title: "Stipa Fecundity analysis"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(lme4)
```

Source data script
```{r data-source}
source('lagged-precip_data-cleaning.R')
```

Starting with a linear mixed-effects model analysis approach. Currently have treatment background categories, want to transition that to actual background densities in those plots. 

Don't have seed estimates for each plot, would need to estimate those from biometric relationships for the ones we do have data for

Avena density in avena_plot tibble
Stipa density can be pulled from stipa_fec tibble probably
```{r lme-analysis}
# phytometers only for now
stipa_fec_phy <- stipa_fec %>% 
  filter(sp_type == "phyto", !is.na(sp_tillers)) %>%
  mutate(seed_est = sp_tillers * seeds_tiller)  # individual-level estimates of total seed count

stipa_count <- stipa_fec %>%
  filter(!(is.na(sp_diam))) %>%
  group_by(block, plot, year) %>%
  summarise(
    sp_bg_count = length(sp_diam) - 1
  )

# combining avena and stipa background densities
stipa_phy <- stipa_fec_phy %>%
  select(year, treatment, block, plot, bg, 
         sp_diam, sp_tillers, seeds_tiller, seed_est) %>%
  left_join(avena_plot, by = join_by(block, plot, year, bg, treatment)) %>%
  select(-c(avena_count, area)) %>%
  left_join(stipa_count, by = join_by(block, plot, year))

m_tillers <- lmer(sp_tillers ~ avena_density + sp_bg_count + 
                    treatment + (1|block) + (1|year),
                  data = stipa_phy)

summary(m_tillers)

stipa_phy %>%
  ggplot(aes(x = avena_density, y = sp_tillers, color = treatment)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  scale_color_manual(values = drought.colors) +
  theme_bw()

stipa_phy %>%
  ggplot(aes(x = sp_bg_count, y = sp_tillers, color = treatment)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  scale_color_manual(values = drought.colors) +
  theme_bw()

stipa_phy %>%
  ggplot(aes(x = avena_density, y = sp_diam, color = treatment)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  scale_color_manual(values = drought.colors) +
  theme_bw()

```

