---
title: "Stipa Fecundity analysis"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(lme4) # linear mixed effects modeling
library(rstan) # direct modeling through Stan
library(tidybayes) # Manipulate Stan objects in a tidy way
library(brms) # Bayesian modeling through Stan
library(broom.mixed) # convernt brms model objects to data frames
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

```

Source data script
```{r data-source}
source('lagged-precip_data-cleaning.R')
```

Starting with a linear mixed-effects model analysis approach. Currently have treatment background categories, want to transition that to actual background densities in those plots. 

Don't have seed estimates for each plot, would need to estimate those from biometric relationships for the ones we do have data for

Avena density in avena_plot tibble
Stipa density can be pulled from stipa_fec tibble probably
```{r lme-analysis}
# phytometers only for now
stipa_fec_phy <- stipa_fec %>% 
  filter(sp_type == "phyto", !is.na(sp_tillers)) %>%
  mutate(seed_est = sp_tillers * seeds_tiller)  # individual-level estimates of total seed count

stipa_count <- stipa_fec %>%
  filter(!(is.na(sp_diam))) %>%
  group_by(block, plot, year) %>%
  summarise(
    sp_bg_count = length(sp_diam) - 1
  )

# combining avena and stipa background densities
stipa_phy <- stipa_fec_phy %>%
  select(year, treatment, block, plot, bg, 
         sp_diam, sp_tillers, seeds_tiller, seed_est) %>%
  left_join(avena_plot, by = join_by(block, plot, year, bg, treatment)) %>%
  select(-c(avena_count, area)) %>%
  left_join(stipa_count, by = join_by(block, plot, year))

# without size (what are treatment and density effects on fecundity, including those via size)
m_tillers <- lmer(sp_tillers ~ avena_density + sp_bg_count + 
                    treatment + (1|block) + (1|year),
                  data = stipa_phy)

summary(m_tillers)

# with size (for a given size, what are treatment and density effects on fecundity?)
# I think this is less useful for our purposes
m_tillers_2 <- lmer(sp_tillers ~ avena_density + sp_bg_count + sp_diam +
                    treatment + (1|block) + (1|year),
                  data = stipa_phy)

summary(m_tillers_2)

# just modeling size
m_size <- lmer(sp_diam ~ avena_density + sp_bg_count + 
                 treatment + (1|block) + (1|year),
               data = stipa_phy)
summary(m_size)
# interestingly, drought treatment has a negative effect on size but a positive effect on fecundity. Negative impacts of avena and stipa competition on fecundity seem to be through size

stipa_phy %>%
  ggplot(aes(x = avena_density, y = sp_tillers, color = treatment)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  scale_color_manual(values = drought.colors) +
  theme_bw()

stipa_phy %>%
  ggplot(aes(x = sp_bg_count, y = sp_tillers, color = treatment)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  scale_color_manual(values = drought.colors) +
  theme_bw()

stipa_phy %>%
  ggplot(aes(x = avena_density, y = sp_diam, color = treatment)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  scale_color_manual(values = drought.colors) +
  theme_bw()

```

# BRMS approach 
```{r stipa-brms}
# preparing dataframe for brms analysis
stipa_phy_2 <- stipa_phy %>% 
  rename(
    stipa_tillers = sp_tillers,
    stipa_bg_count = sp_bg_count,
    stipa_diam = sp_diam
  ) %>%
  filter(bg %in% c("sp_low", "sp_high", "a_low", "a_high", "control")) %>%
  select(c(year, treatment, block, plot, bg, stipa_diam, stipa_tillers, 
           avena_density, stipa_bg_count)) %>%
  mutate(avena_dens_2 = avena_density/100,
         year = as.factor(year))

priors <- c(set_prior("normal(0,1)", class = "b", coef = "treatmentdrought"),
            set_prior("normal(0,0.1)", class = "b", coef = "avena_dens_2"),
            set_prior("normal(0,0.1)", class = "b", coef = "stipa_bg_count"))

b_tillers <- brm(stipa_tillers ~ treatment * avena_dens_2 + treatment * stipa_bg_count + treatment * year +
                    (1|block),
                 family = poisson(),
                 data = stipa_phy_2, 
                 prior = priors, 
                 control = list(adapt_delta = 0.97, max_treedepth = 12),
                 iter = 3000, warmup = 1000,
                 init = 0)
rhat(b_tillers)

# pairs(b_tillers)  # check sampling structure

# prior_summary(b_tillers)  # check that priors are the same way as planned

b_tidy <- tidy(b_tillers)
summary(b_tillers)

pp_check(b_tillers)

plot(conditional_effects(b_tillers), ask = FALSE)
```


# Stan approach -- modeling stipa fecundity specifically
Starting by just modeling stipa fecundity as a function of treatment, background, with random effects for block and year
Matching structure of m_tillers above
```{r stipa-stan}

```

