---
title: "Lagged Precip Biomass"
output: html_document
date: "2025-10-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(lme4)

theme_cw <- function () { 
  theme_bw(base_size=12) %+replace% 
    theme(
      panel.background = element_blank(), 
      plot.background = element_blank(), 
      axis.ticks = element_line(colour = "grey70", size = rel(0.5)),
      panel.grid.minor = element_blank(), 
      panel.grid.major.x = element_blank(),
      legend.background = element_blank(), 
      legend.key = element_blank(),
      strip.background = element_blank(), 
     # axis.text=element_text(size=12),
     # strip.text=element_text(size=12),
      complete = TRUE
      )
}
```

## Reading in Data


```{r data-read}
datpath <- "data/"

# block and plot key
trt.key <- read.csv(paste0(datpath, "plot-layout/block-key.csv")) %>%
  select(block:treatment) %>%
  mutate(treatment = ifelse(treatment == "dry", "drought", "ambient"),
         block = as.factor(block))

# Biomass 2024
biomass <- read.csv(paste0(datpath, "lagged-precip_stipa-avena-biomass_2024.csv")) %>%
  left_join(trt.key, by = "block") %>%
  rename("gross.biomass" = "biomass..incl..bag.",
         "bag.mass" = "mass.of.bag..g.") %>%
  mutate(biomass = gross.biomass - bag.mass,
         sample = str_remove_all(sample, " "), # there might be a cleaner way to strip whitespace
         bg = case_when(
           bg == "" ~ "outside plot",
           .default = bg
         )) 


drought.colors <- c("#0077CC", "#CD622E")

```

There's a couple ways to think about biomass. We can think about total biomass in each plot, with the combination of stipa and avena. Or we can think about how each of those differs with drought treatment and in the presence of the other

# Biomass preliminary
Stipa phytometer biomass is similar or slightly higher in drought treatment, highest in control, med in avena, lowest in sp

Avena biomass is lower in drought treatments

Stipa background biomass doesn't differ much with drought
```{r stipa}
biomass_st_phy <- biomass %>% filter(sample == "stipaphytometer")

biomass_st_phy %>% 
  ggplot(aes(x = bg, y = biomass, color = treatment)) +
  geom_boxplot() + 
  scale_color_manual(values = drought.colors) +
  theme_cw() +
  ylab('Stipa phytometer biomass')


biomass_avena <- biomass %>% filter(sample == "avena")

biomass_avena %>%
  ggplot(aes(x = bg, y = biomass, color = treatment)) +
  geom_boxplot() + 
  scale_color_manual(values = drought.colors) +
  theme_cw() +
  ylab('Avena biomass')

biomass_st_bg <- biomass %>% filter(sample == "stipabackground")

biomass_st_bg %>%
  ggplot(aes(x = bg, y = biomass, color = treatment)) +
  geom_boxplot() + 
  scale_color_manual(values = drought.colors) +
  theme_cw() +
  ylab('Stipa background biomass')

```


