---
title: "Prelim Analyses 2023"
output: html_document
date: "2024-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)

theme_cw <- function () { 
  theme_bw(base_size=12) %+replace% 
    theme(
      panel.background = element_blank(), 
      plot.background = element_blank(), 
      axis.ticks = element_line(colour = "grey70", size = rel(0.5)),
      panel.grid.minor = element_blank(), 
      panel.grid.major.x = element_blank(),
      legend.background = element_blank(), 
      legend.key = element_blank(),
      strip.background = element_blank(), 
     # axis.text=element_text(size=12),
     # strip.text=element_text(size=12),
      complete = TRUE
      )
}
```

## Reading in Data


```{r data-read}
datpath <- "data/"

# block and plot key
trt.key <- read.csv(paste0(datpath, "plot-layout/block-key.csv")) %>%
  select(block:treatment) %>%
  mutate(treatment = ifelse(treatment == "dry", "drought", "ambient"))

# Stipa fecundity 2023
stipa_fec_23 <- read.csv(paste0(datpath,"lagged-precip_stipa-fecundity_2023-04.csv")) %>%
  rename(stipa_type = Stipa..type,
         stipa_diam = Stipa.diam..cm.,
         weed_gram = X..weed.gram, 
         weed_forb = X..weed.forb) %>%
  rename_with(.,~ tolower(gsub( ".", "_", .x, fixed = TRUE))) %>%
  rename_with(.,~ gsub( "stipa", "sp", .x, fixed = TRUE)) %>%
  mutate(year = 2023) 

# Stipa fecundity 2024
stipa_fec_24 <- read.csv(paste0(datpath,"lagged-precip_stipa-fecundity_2024-04.csv")) %>%
  rename_with(.,~ tolower(gsub( ".", "_", .x, fixed = TRUE))) %>%
  mutate(year = 2024,
         sp_seeds_per_tiller = as.numeric(sp_seeds_per_tiller)) 

# Join 2023 and 2024 stipa fecundity
stipa_fec <- full_join(stipa_fec_23, stipa_fec_24, 
                       by = join_by(block, plot, bg, sp_type,
                                    sp_diam == sp_diam_1, sp_tillers, 
                                    seeds_tiller == sp_seeds_per_tiller, year,
                                    notes)) %>%
  left_join(trt.key, by = "block") %>%
  mutate(bg = case_when(  
    bg == "S_P low" | bg == "S_Plow" ~ "sp_low",
    bg == "S_P high" | bg == "S_Phigh" ~ "sp_high",
    bg == "S_s low" | bg == "S_slow" ~ "ss_low",
    bg == "S_s high" | bg == "S_shigh" ~ "ss_high",
    bg == "A low" | bg == "A_low" ~ "a_low",
    bg == "A high" | bg == "A_high" ~ "a_high",
    bg == "control" ~ "control",
    .default = "NA"
  )) # making plot background labels consistent



# Avena density and fecundity 
# NOTE: background plot labels are not consistent across years here -- need to update so they are

# Avena 2023
# data was entered in a weird wide format, took a little reshuffling to get it into a workable long format
avena_23 <-  read.csv(paste0(datpath,"lagged-precip_avena_2023-04.csv")) %>%
  rename_with(.,~ tolower(gsub( ".", "_", .x, fixed = TRUE))) %>%
  mutate(across(sub1_plants:full_im_tillers, as.numeric),
         year = 2023) %>%
  pivot_longer(cols = sub1_plants:full_im_tillers) %>%
  separate_wider_delim(cols = name, delim = "_", names = c("location", "info"), too_many = "merge") %>%
  pivot_wider(names_from = "info", values_from = "value") %>%
  pivot_longer(cols = seeds_1:im_5) %>%
  separate_wider_delim(cols = name, delim = "_", names = c("tiller_type", "rep")) %>%
  pivot_wider(names_from = "tiller_type", values_from ="value") %>%
  rename(im_seeds = im)

# Avena 2024
avena_24 <-  read.csv(paste0(datpath,"lagged-precip_avena_2024-05.csv")) %>%
  rename_with(.,~ tolower(gsub( ".", "_", .x, fixed = TRUE))) %>%
  mutate(year = 2024, 
         location = case_when(
           location == "" ~ "full",  # missing location info in one row
           .default = location))

avena <- full_join(avena_23, avena_24, by = join_by(block, plot, bg, location, 
                                                    plants, seeds, im_tillers, year)) %>%
  left_join(trt.key, by = "block") %>%
  mutate(bg = case_when(  
    bg == "S_P low" | bg == "S_Plow" ~ "sp_low",
    bg == "S_P high" | bg == "S_Phigh" ~ "sp_high",
    bg == "S_s low" | bg == "S_slow" ~ "ss_low",
    bg == "S_s high" | bg == "S_shigh" ~ "ss_high",
    bg == "A low" | bg == "A_low" ~ "a_low",
    bg == "A high" | bg == "A_high" ~ "a_high",
    bg == "control" ~ "control",
    .default = "NA"
  )) # making plot background labels consistent


## survival/recruitment

# 2023, need to double-check what the sections of 1, 2, and 3 are. 2/3 of the plot was overseeded
# in the second year, so this is separating out first added vs newly added seed
# but need to check which side 1 vs 3 are

# this is also the dataset that has weed info for this time period
ss_23 <- read.csv(paste0(datpath,"lagged-precip_survival_2023-02.csv")) %>%
  mutate(ss_old_tot = ss.old.1 + ss.old.2 + ss.old.3, 
         ss_new_tot = ss.new.1 + ss.new.2 + ss.new.3) %>% 
  left_join(trt.key, by = "block") %>%
  mutate(bg = case_when(  
    bg == "S_P low" | bg == "S_Plow" ~ "sp_low",
    bg == "S_P high" | bg == "S_Phigh" ~ "sp_high",
    bg == "S_s low" | bg == "S_slow" ~ "ss_low",
    bg == "S_s high" | bg == "S_shigh" ~ "ss_high",
    bg == "A low" | bg == "A_low" ~ "a_low",
    bg == "A high" | bg == "A_high" ~ "a_high",
    bg == "control" ~ "control",
    .default = "NA"
  ))

# 2024 data needs cleaning of block/background names
# also not necessarily directly connected to 2023 data, need to think about how best to combine

drought.colors <- c("#0077CC", "#CD622E")

```

## Stipa fecundity

Stipa adult fecundity, spring 2023

Remember that plots 9, 10, 11, and 12 don't have a phytometer (9 and 10 were extra stipa high-density background plots, 11 and 12 were the added ss and control plots which generally didn't work out)

Dropped block 9 in 2023 and additionally dropped block 3 in 2024, and some additional plots within blocks (usually 9 and 10 when gopher damage was too high)


For now, only focusing on phytometers -- note that we also have fecundity information from low- and high-density stipa background plots, which can increase the information when incorporated in a model

Analysis questions:
- Focus on phyotometers only, or include fecundity and surivival information from low- and high-density stipa background plots? 

- How to incorporate seeds/tiller information? 
```{r stipa-fecundity}

# phytometers only for now
stipa_fec_phy <- stipa_fec %>% filter(sp_type == "phyto", !is.na(sp_tillers))

# number of tillers
stipa_fec_phy %>% 
  ggplot(aes(x = bg, y = sp_tillers, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

# seeds per tiller
stipa_fec_phy %>% 
  ggplot(aes(x = bg, y = seeds_tiller, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()
# size
stipa_fec_phy %>% 
  ggplot(aes(x = bg, y = sp_diam, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()


# large stipa in background plots
stipa_fec_big <- stipa_fec %>% filter(sp_type == "big", !is.na(sp_tillers))

# number of tillers
stipa_fec_big %>% 
  ggplot(aes(x = bg, y = sp_tillers, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

# seeds per tiller
stipa_fec_big %>% 
  ggplot(aes(x = bg, y = seeds_tiller, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

# size
stipa_fec_big %>% 
  ggplot(aes(x = bg, y = sp_diam, color = treatment)) +
  facet_wrap(vars(year)) +
  scale_color_manual(values = drought.colors) +
  geom_boxplot()

# tillers by size in control plots
stipa_fec %>%
  filter(bg == "control", sp_type %in% c("phyto", "big", "med")) %>%
  ggplot(aes(x = sp_diam, y = sp_tillers, color = treatment, shape = sp_type)) +
  facet_wrap(vars(year)) + 
  scale_color_manual(values = drought.colors) +
  geom_point() +
  geom_smooth(method = "lm")

# tillers by size in SP plots
stipa_fec %>%
  filter(bg %in% c("sp_high", "sp_low"), sp_type %in% c("phyto", "big", "med")) %>%
  ggplot(aes(x = sp_diam, y = sp_tillers, 
             color = treatment, shape = sp_type, linetype = sp_type)) +
  facet_wrap(vars(year)) + 
  scale_color_manual(values = drought.colors) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)



```

## Avena Density and Fecundity

```{r avena}

# avena density calculation
avena <- avena %>%
  mutate(
    density = case_when(
      location == "full" ~ plants * 2, # 0.5 x 1m full plot
      .default = plants * 25 # 0.2 x 0.2 m subplot
    )
  )  

# want to also then average densities across subplots so we only have one average plant density value per plot
# but need to do this carefully so it's a real average not counting per number of plants measured

# avena density preliminary graph
avena %>% 
  ggplot(aes(x = bg, y = density, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

# avena fecundity preliminary graph (seeds/plant way lower in 2024 -- were we counting the same way? this does fit with my memory)
avena %>% 
  ggplot(aes(x = bg, y = seeds, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

```



## Stipa seedling density, survival, and fecundity
```{r stipa-seedling}

ss_23 %>%
  ggplot(aes(x = bg, y = ss_old_tot, color = treatment)) +
  scale_color_manual(values = drought.colors) +
  geom_boxplot()

ss_23 %>%
  ggplot(aes(x = bg, y = ss_new_tot, color = treatment)) +
  scale_color_manual(values = drought.colors) +
  geom_boxplot()

```

