---
title: "Prelim Analyses 2023"
output: html_document
date: "2024-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(lme4)

theme_cw <- function () { 
  theme_bw(base_size=12) %+replace% 
    theme(
      panel.background = element_blank(), 
      plot.background = element_blank(), 
      axis.ticks = element_line(colour = "grey70", size = rel(0.5)),
      panel.grid.minor = element_blank(), 
      panel.grid.major.x = element_blank(),
      legend.background = element_blank(), 
      legend.key = element_blank(),
      strip.background = element_blank(), 
     # axis.text=element_text(size=12),
     # strip.text=element_text(size=12),
      complete = TRUE
      )
}
```

## Reading in Data


```{r data-read}
datpath <- "data/"

# block and plot key
trt.key <- read.csv(paste0(datpath, "plot-layout/block-key.csv")) %>%
  select(block:treatment) %>%
  mutate(treatment = ifelse(treatment == "dry", "drought", "ambient"))

# Stipa fecundity 2023
stipa_fec_23 <- read.csv(paste0(datpath,"lagged-precip_stipa-fecundity_2023-04.csv")) %>%
  rename(stipa_type = Stipa..type,
         stipa_diam = Stipa.diam..cm.,
         weed_gram = X..weed.gram, 
         weed_forb = X..weed.forb) %>%
  rename_with(.,~ tolower(gsub( ".", "_", .x, fixed = TRUE))) %>%
  rename_with(.,~ gsub( "stipa", "sp", .x, fixed = TRUE)) %>%
  mutate(year = 2023) 

# Stipa fecundity 2024
stipa_fec_24 <- read.csv(paste0(datpath,"lagged-precip_stipa-fecundity_2024-04.csv")) %>%
  rename_with(.,~ tolower(gsub( ".", "_", .x, fixed = TRUE))) %>%
  mutate(year = 2024,
         sp_seeds_per_tiller = as.numeric(sp_seeds_per_tiller)) 

# Join 2023 and 2024 stipa fecundity
stipa_fec <- full_join(stipa_fec_23, stipa_fec_24, 
                       by = join_by(block, plot, bg, sp_type,
                                    sp_diam == sp_diam_1, sp_tillers, 
                                    seeds_tiller == sp_seeds_per_tiller, year,
                                    notes)) %>%
  left_join(trt.key, by = "block") %>%
  mutate(bg = case_when(  
    bg == "S_P low" | bg == "S_Plow" ~ "sp_low",
    bg == "S_P high" | bg == "S_Phigh" ~ "sp_high",
    bg == "S_s low" | bg == "S_slow" ~ "ss_low",
    bg == "S_s high" | bg == "S_shigh" ~ "ss_high",
    bg == "A low" | bg == "A_low" ~ "a_low",
    bg == "A high" | bg == "A_high" ~ "a_high",
    bg == "control" ~ "control",
    .default = "NA"
  )) # making plot background labels consistent



# Avena density and fecundity 
# NOTE: background plot labels are not consistent across years here -- need to update so they are

# Avena 2023
# data was entered in a weird wide format, took a little reshuffling to get it into a workable long format
avena_23 <-  read.csv(paste0(datpath,"lagged-precip_avena_2023-04.csv")) %>%
  rename_with(.,~ tolower(gsub( ".", "_", .x, fixed = TRUE))) %>%
  mutate(across(sub1_plants:full_im_tillers, as.numeric),
         year = 2023) %>%
  pivot_longer(cols = sub1_plants:full_im_tillers) %>%
  separate_wider_delim(cols = name, delim = "_", names = c("location", "info"), too_many = "merge") %>%
  pivot_wider(names_from = "info", values_from = "value") %>%
  pivot_longer(cols = seeds_1:im_5) %>%
  separate_wider_delim(cols = name, delim = "_", names = c("tiller_type", "rep")) %>%
  pivot_wider(names_from = "tiller_type", values_from ="value") %>%
  rename(im_seeds = im)

# Avena 2024
avena_24 <-  read.csv(paste0(datpath,"lagged-precip_avena_2024-05.csv")) %>%
  rename_with(.,~ tolower(gsub( ".", "_", .x, fixed = TRUE))) %>%
  mutate(year = 2024, 
         location = case_when(
           location == "" ~ "full",  # missing location info in one row
           .default = location))

avena <- full_join(avena_23, avena_24, by = join_by(block, plot, bg, location, 
                                                    plants, seeds, im_tillers, year)) %>%
  left_join(trt.key, by = "block") %>%
  mutate(bg = case_when(  
    bg == "S_P low" | bg == "S_Plow" ~ "sp_low",
    bg == "S_P high" | bg == "S_Phigh" ~ "sp_high",
    bg == "S_s low" | bg == "S_slow" ~ "ss_low",
    bg == "S_s high" | bg == "S_shigh" ~ "ss_high",
    bg == "A low" | bg == "A_low" ~ "a_low",
    bg == "A high" | bg == "A_high" ~ "a_high",
    bg == "control" ~ "control",
    .default = "NA"
  )) # making plot background labels consistent


## survival/recruitment

# 2023, need to double-check what the sections of 1, 2, and 3 are. 2/3 of the plot was overseeded
# in the second year, so this is separating out first added vs newly added seed
# but need to check which side 1 vs 3 are

# this is also the dataset that has weed info for this time period
ss_23 <- read.csv(paste0(datpath,"lagged-precip_survival_2023-02.csv")) %>%
  mutate(ss_old_tot = ss.old.1 + ss.old.2 + ss.old.3, 
         ss_new_tot = ss.new.1 + ss.new.2 + ss.new.3) %>% 
  left_join(trt.key, by = "block") %>%
  mutate(bg = case_when(  
    bg == "S_P low" | bg == "S_Plow" ~ "sp_low",
    bg == "S_P high" | bg == "S_Phigh" ~ "sp_high",
    bg == "S_s low" | bg == "S_slow" ~ "ss_low",
    bg == "S_s high" | bg == "S_shigh" ~ "ss_high",
    bg == "A low" | bg == "A_low" ~ "a_low",
    bg == "A high" | bg == "A_high" ~ "a_high",
    bg == "control" ~ "control",
    .default = "NA"
  ))

# 2024 data needs cleaning of block/background names
# also not necessarily directly connected to 2023 data, need to think about how best to combine

drought.colors <- c("#0077CC", "#CD622E")

```

## Stipa fecundity

Stipa adult fecundity, spring 2023

Remember that plots 9, 10, 11, and 12 don't have a phytometer (9 and 10 were extra stipa high-density background plots, 11 and 12 were the added ss and control plots which generally didn't work out)

Dropped block 9 in 2023 and additionally dropped block 3 in 2024, and some additional plots within blocks (usually 9 and 10 when gopher damage was too high)


For now, only focusing on phytometers -- note that we also have fecundity information from low- and high-density stipa background plots, which can increase the information when incorporated in a model

Stipa phytometers are largest in control plots, slightly smaller in avena and ss plots,and smallest in SP plots, with corresponding difference in fecundity as size seems to be the biggest predictor of fecundity. Drought treatment doesn't consistently affect stipa size or fecundity, but may somewhat increase number of seeds per tiller. There may be a negative impact of drought in the Avena background plots, which would be really interesting

Analysis questions:
- Focus on phyotometers only, or include fecundity and surivival information from low- and high-density stipa background plots? 

- How to incorporate seeds/tiller information? 

- Do we have enough survival information to look at drought effects, or is it mostly just gopher damage?
```{r stipa-fecundity}

# phytometers only for now
stipa_fec_phy <- stipa_fec %>% filter(sp_type == "phyto", !is.na(sp_tillers))

# number of tillers
stipa_fec_phy %>% 
  ggplot(aes(x = bg, y = sp_tillers, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

# seeds per tiller
stipa_fec_phy %>% 
  ggplot(aes(x = bg, y = seeds_tiller, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()
# size
stipa_fec_phy %>% 
  ggplot(aes(x = bg, y = sp_diam, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

# tillers by size by treatment
stipa_fec_phy %>%
  mutate(
    bg.2 = case_when(
      bg %in% c("a_low", "a_high") ~ "avena",
      bg %in% c("sp_low", "sp_high") ~ "sp",
      bg %in% c("ss_low", "ss_high") ~ "ss",
      bg == "control" ~ "control",
      .default = "NA"
    )) %>%
  ggplot(aes(x = sp_diam, y = sp_tillers, 
             color = bg.2, shape = treatment, linetype = treatment)) +
  facet_wrap(vars(year)) + 
#  scale_color_manual(values = drought.colors) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_cw()
  

stipa_phy_size_m <- lmer(sp_diam ~ bg + treatment + year + (1|block), 
                    data = stipa_fec_phy)
summary(stipa_phy_size_m)

stipa_phy_fec_m <- lmer(sp_tillers ~ bg + sp_diam + treatment + year + (1|block), 
                    data = stipa_fec_phy)
summary(stipa_phy_fec_m)

stipa_phy_fec2_m <- lmer(seeds_tiller ~ bg + treatment + year + (1|block), 
                    data = stipa_fec_phy)
summary(stipa_phy_fec2_m)

# large stipa in background plots
stipa_fec_big <- stipa_fec %>% filter(sp_type == "big", !is.na(sp_tillers))

# number of tillers
stipa_fec_big %>% 
  ggplot(aes(x = bg, y = sp_tillers, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

# seeds per tiller
stipa_fec_big %>% 
  ggplot(aes(x = bg, y = seeds_tiller, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

# size
stipa_fec_big %>% 
  ggplot(aes(x = bg, y = sp_diam, color = treatment)) +
  facet_wrap(vars(year)) +
  scale_color_manual(values = drought.colors) +
  geom_boxplot()

# tillers by size in control plots
stipa_fec %>%
  filter(bg == "control", sp_type %in% c("phyto", "big", "med")) %>%
  ggplot(aes(x = sp_diam, y = sp_tillers, color = treatment, shape = sp_type)) +
  facet_wrap(vars(year)) + 
  scale_color_manual(values = drought.colors) +
  geom_point() +
  geom_smooth(method = "lm")

# tillers by size in SP plots
stipa_fec %>%
  filter(bg %in% c("sp_high", "sp_low"), sp_type %in% c("phyto", "big", "med")) %>%
  ggplot(aes(x = sp_diam, y = sp_tillers, 
             color = treatment)) +
  facet_wrap(vars(year)) + 
  scale_color_manual(values = drought.colors) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE)

# drought has a lower intercept but a higher slope
# these are very minor differences compared to the effect of size
stipa_fec$year <- as.factor(stipa_fec$year)
stipa_fec_m <- lmer(sp_tillers ~ sp_diam*treatment*year + (1|block.id), 
                  data = stipa_fec)

summary(stipa_fec_m)

stipa_size_m <- lmer(sp_diam ~ treatment*year + (1|block.id), data = stipa_fec)
summary(stipa_size_m)

```

## Avena Density and Fecundity
Drought has minimal effects on density in 2023, neg in 2024 (I think I'm interpreting that right, need to check). Drought has minimal effects on avena fecundity

Overall avena density was higher in 2024 (the less intensely wet year) -- although that may be partially due to the previous year's seed drop. When we compare the seeded side (sub1) to the unseeded site (sub2) they are quite similar

```{r avena}

# avena density calculation
avena <- avena %>%
  mutate(
    density = case_when(
      location == "full" ~ plants * 2.18, # 0.5 x 1m full plot, minus the two .2 x .2 subplots
      .default = plants * 25 # 0.2 x 0.2 m subplot
    )
  )

avena %>%
  filter(year == 2024) %>%
  ggplot(aes(x = location, y = density, color = treatment)) + 
  facet_wrap(vars(bg), scales = "free") +
  scale_color_manual(values = drought.colors) +
  geom_boxplot()

# might be better to add the subplot densities back into the full plot and then take the averages?
avena.plot <- avena %>%
  group_by(block, plot, bg, year, treatment) %>%
  summarise(
    mean.dens = mean(density, na.rm = TRUE) 
  )

# want to also then average densities across subplots so we only have one average plant density value per plot
# but need to do this carefully so it's a real average not counting per number of plants measured

# avena density preliminary graph
avena.plot %>% 
  ggplot(aes(x = bg, y = mean.dens, color = treatment)) +
  facet_wrap(vars(year)) +
  scale_color_manual(values = drought.colors) +
  geom_boxplot()

avena.plot$year = as.factor(avena.plot$year)
avena_m <- lmer(mean.dens ~ bg + year*treatment + (1|block), 
                data = avena.plot)
summary(avena_m)

# avena fecundity preliminary graph (seeds/plant way lower in 2024 -- were we counting the same way? this does fit with my memory)
avena %>% 
  ggplot(aes(x = bg, y = seeds, color = treatment)) +
  facet_wrap(vars(year)) +
    scale_color_manual(values = drought.colors) +
  geom_boxplot()

avena_fec_m <- lmer(seeds ~ bg + year*treatment + (1|block), 
                data = avena)
summary(avena_fec_m)

```



## Stipa seedling density, survival, and fecundity
Stipa seedlings emergence/survival may be slightly lower in drought treatments
Still need to look at stipa seedling survival estimates overall
```{r stipa-seedling}

ss_23 %>%
  ggplot(aes(x = bg, y = ss_old_tot, color = treatment)) +
  scale_color_manual(values = drought.colors) +
  geom_boxplot()

ss_23 %>%
  ggplot(aes(x = bg, y = ss_new_tot, color = treatment)) +
  scale_color_manual(values = drought.colors) +
  geom_boxplot()

```

# Next steps
I think the next stage is combining things together into a BH model with alpha competition estimates. I'm not entirely sure how to break down size/fecundity vs population size (which we don't really have) but maybe a separate model for SP size and/or fecundity with predictors of actual SP density and AV density and drought treatment (rather than just the categorical background treatment) -- not BH, just a mixed model? 

For Avena, probably need to wade back in to density methods to get a sense of seeds in vs seeds out (if that's something we decide we're incoporating into the model)

Can also get into the biomass data